Specialized Cost Function for Maximizing Body Area Networks Lifetime Using Global Routing Algorithms

### 2011-09-08 docs/thesis/paper.tex
Added changes suggested by Dr. Tsouri. Still needs tons of work.

\section{Introduction}
Wireless Body Area Networks (WBANs) consist of several wireless sensors located around a human body. These sensors may measure several biological signals, movement, and temperature. Due to major improvements in power consumption and constantly shrinking devices, WBANs are becoming more widely used. Due to the small form factor of these devices, the battery size is limited. While the sensors themselves may be extremely power efficient, all of the measured data must be transmitted over a much less efficient wireless link. One benefit of WBANs is that they rarely include more than a dozen wireless devices over a small area. This constraint allows for the use of routing techniques not suitable for larger wireless networks.

There are several routing protocols designed specifically for WBANs. Braem et al. developed the CICADA protocol, which consists of a spanning tree architecture with a time-division scheme for transmission scheduling~\cite{protocol:CICADA}. The primary downside to this method is that nodes closer to the root will deplete their energy source faster due to the need to relay messages from children nodes. Other protocols focus on the reliability of the network. Constant network changes are a problem with WBANs. If sensors are placed in the extremities, the network will change as the body changes positions. Quwaider et al. developed a protocol tolerant to network changes~\cite{routing:storeandforward}. They propose a store-and-forward method that maximizes the likelihood of a packet reaching its destination. Each packet is stored by multiple devices and retransmitted, which consumes more power. One solution for this problem was proposed by Ehyaie et al.~\cite{relay:networklife}, it consists of using dedicated, non-sensing, relay nodes with larger power sources. While this method increases the network lifetime, it requires more hardware. Finally, Nabi et al. propose a similar store-and-forward method, but integrates Transmit Power Adaptation (TPA)~\cite{relay:transmitpoweradaptation}. Nodes keep track of neighbors and use power control to use the smallest transmit power while maintaining a specific link quality.

@@

Each device's energy use is accumulated as shown in Equation~\ref{eqn:accumulated_energy}, where $j$ denotes the device, $i$ is the current polling round, $\alpha$ is the channel attenuation, and $RSSI^T$ is the target RSSI. The current energy used is incremented by the energy used to transmit a single packet while maintaining $RSSI^T$. The channel attenuation for the selected link,$\alpha$, can be seen in Equation~\ref{eqn:alpha}, where RSSI is the received power and $P_tx$ is the transmitted power. The actual link cost is computed by calculating the energy used by a device \emph{if} that link is selected and subtracting the average energy used across all devices. Equation~\ref{eqn:link_cost} is normalized by the current polling round $i$.
\begin{equation}
\label{eqn:alpha}
\alpha = \frac{RSSI}{P_{tx}}
\end{equation}
\begin{equation}
\label{eqn:accumulated_energy}
E^j_{acc_{(i)}} = E^j_{acc_{(i-1)}} + \frac{RSSI^T}{\alpha}
\end{equation}
\begin{equation}
\label{eqn:link_cost}
C^j_{l,k} = \left[ \left(\frac{ E^l_{acc_{(i-1)}} + \frac{RSSI^T}{\alpha_{l,k}}}{i}\right) - \frac{\sum_jE^l_{acc_{(i-1)}}}{i-1} \right]
\end{equation}
$\ll$describe network (centralized network with powerful AP and multiple sensors able to act as relays, power control), figure 1 (like in paper fig.1 + symbols for channel attenuation, cost, sensor ID,…) and describe symbols in figure.$\gg$
$\ll$Define accumulated energy (normalized power), cost function and all relevant symbols$\gg$
$\ll$rational for cost function$\gg$

\section{Performance Analysis}
$\ll$”in what follows we …. (state simulation and hardware), implementation of cost function in Dijktra (ref),step wide pseudo code for algorithm, comparison approach to “normal” cost function with Dijkstra$\gg$
$\ll$TI platform (in detail with calibration – all data required to replicate) for RSSI and Disjkstra$\gg$

@@

\section{Results}
$\ll$describe figure  3 and interpret results – network lifetime (like fig.8 in paper with Jeff)$\gg$ 
$\ll$describe figure  4 and interpret results – energy per bit (like fig.9 in paper with Jeff)$\gg$ 
$\ll$* maybe use CDFs?...$\gg$

>>>>

\section{Introduction}
Wireless Body Area Networks (WBANs) consist of several wireless sensors located around a human body. These sensors may measure several biological signals, movement, and temperature. Due to major improvements in power consumption and constantly shrinking devices, WBANs are becoming ubiquitous. Due to the small form factor of these devices, the battery size is limited. While the sensors themselves may be extremely power efficient, all of the measured data must be transmitted over a much less efficient wireless link. One benefit of WBANs is that they rarely include more than a dozen wireless devices over a small area. This constraint allows for the use of routing techniques not suitable for larger wireless sensor networks.

There are several routing protocols designed specifically for WBANs. Braem et al. developed the CICADA protocol, which consists of a spanning tree architecture with a time-division scheme for transmission scheduling~\cite{protocol:CICADA}. The primary downside to this method is that nodes closer to the root will deplete their energy source faster due to the need to relay messages from children nodes. Other protocols focus on the reliability of the network$\ll$citation needed$\gg$. Dynamic networks are a problem with WBANs. If sensors are placed in the extremities, the network will change as the body changes positions. Quwaider et al. developed a protocol tolerant to network changes~\cite{routing:storeandforward}. They propose a store-and-forward method that maximizes the likelihood of a packet reaching its destination. Each packet is stored by multiple devices and retransmitted, which consumes more power. One solution for this problem was proposed by Ehyaie et al.~\cite{relay:networklife}, it consists of using dedicated, non-sensing, relay nodes with larger power sources. While this method increases the network lifetime$\ll$define term$\gg$, it requires more hardware. Finally, Nabi et al. propose a similar store-and-forward method, but integrates Transmit Power Adaptation (TPA)~\cite{relay:transmitpoweradaptation}. Nodes keep track of neighbors and use power control to use the smallest transmit power while maintaining a specific link quality.

@@

Each device's energy use is accumulated as shown in Equation~\ref{eqn:accumulated_energy}, where $j$ denotes the device ID, $i$ is the current polling round, $\alpha$ is the channel attenuation, and $RSSI^T$ is the target RSSI. The current energy used is incremented by the energy used to transmit a single packet while maintaining $RSSI^T$. The channel attenuation for the selected link,$\alpha$, can be seen in Equation~\ref{eqn:alpha}, where RSSI is the received power and $P_tx$ is the transmitted power. The actual link cost is computed by calculating the energy used by a device \emph{if} that link is selected and subtracting the average energy used across all devices. Equation~\ref{eqn:link_cost} is normalized by the current polling round $i$.
\begin{equation}
\label{eqn:alpha}
\alpha = \frac{RSSI}{P_{tx}}
\end{equation}
\begin{equation}
\label{eqn:accumulated_energy}
E^j_{acc_{(i)}} = E^j_{acc_{(i-1)}} + \frac{RSSI^T}{\alpha}
\end{equation}
\begin{equation}
\label{eqn:link_cost}
C^j_{l,k} = \left[ \left(\frac{ E^l_{acc_{(i-1)}} + \frac{RSSI^T}{\alpha_{l,k}}}{i}\right) - \frac{\sum_jE^l_{acc_{(i-1)}}}{i-1} \right]
\end{equation}

Once all link costs are calculated, Dijkstra's algorithm~\cite{dijkstra:algorithm} is used to compute the shortest path.

This algorithm works under several assumptions. The first is that this is a centralized network with one central Access Point (AP) with virtually unlimited energy and  computing power. The second is that each wireless sensor, or End Device(ED), can act as a relay for messages from other nodes. The Third and last requirement is that the link quality,or channel attenuation, between all EDs and AP is known. All routing calculations take place in the AP.

$\ll$describe network (centralized network with powerful AP and multiple sensors able to act as relays, power control), figure 1 (like in paper fig.1 + symbols for channel attenuation, cost, sensor ID,…) and describe symbols in figure.$\gg$

$\ll$Define accumulated energy (normalized power), cost function and all relevant symbols$\gg$

$\ll$rational for cost function$\gg$

\section{Performance Analysis}
$\ll$”in what follows we …. (state simulation and hardware), implementation of cost function in Dijktra (ref),step wide pseudo code for algorithm, comparison approach to “normal” cost function with Dijkstra$\gg$

$\ll$TI platform (in detail with calibration – all data required to replicate) for RSSI and Disjkstra$\gg$

@@

\section{Results}

$\ll$describe figure  3 and interpret results – network lifetime (like fig.8 in paper with Jeff)$\gg$

$\ll$describe figure  4 and interpret results – energy per bit (like fig.9 in paper with Jeff)$\gg$

$\ll$* maybe use CDFs?...$\gg$
